World Heritage Sites on Wikipedia: Controversies in a Context of Constrained Agency for Cultural Heritage Activism

### 2020-11-22 analysis/paper/paper.Rmd
finished editing the text

The highly detailed edit histories that Wikipedia keeps for every article allow us to further investigate spatial and temporal questions relating to engagement with the past and conflicts surrounding CS-WHL sites. For example, when an article is edited by a user who does not have a Wikipedia user account (or is not logged into their account), their edit is identified by that person's IP address. This IP address can be used to geo-locate the user to the country they were in when the made the edit. We geo-located all edits with IP addresses for all Wikipedia articles CS-WHL sites to determine the country of origin of those edits. This helps us to answer the question: are the editors of articles about CS-WHL located near the sites they edit, indicating local community interest in the online representation of their heritage? Or are editors located in other countries, implying a digital colonialism of world heritage information on Wikipedia where one community's heritage is interpreted and communicated by another? The time and date stamps attached to every edit on every article allows us to investigate temporal patterns of activity on CS-WHL Wikipedia articles. Analyses of these data help us to answer the question: is Wikipedia editing activity correlated with events outside of Wikipedia relating to the CS-WHL sites, such as conflict events, or their inscription on the WHL?

@@

# Reproducibility and Open Source Materials for This Study

@@

Our results establish Wikipedia as site and community demonstrating active engagement with cultural heritage [@pentzold2017digging], especially cultural sites on the UNESCO World Heritage List. We have defined Wikipedia as an online peer production community with technical and cultural qualities that separate it from social media such as Twitter and Facebook. Editing mechanics, bots, reverts, and talk pages work together to enable and limit discourse and debate about CS-WHL on Wikipedia. The goal-driven culture of Wikipedia editors, and the organisation's policies that guide the production of articles strongly constrain heritage discourse, such that the exchange of different perspectives are rarely found in articles themselves, but buried in technical and highly specific discussions on article talk pages. Our results show how these distinctive qualities generate a technicity that shapes how people engage with elements of the past and attribute social and cultural meanings on Wikipedia. These resemble the "contingent collaborations" and "productive frictions" described by @tsing2011friction where local, individual actions of knowledge creation are circumscribed by forces and structures that encompass their expression, but also give meaning to their interactions.

We found that many of the Wikipedia articles about CS-WHL are central to the encyclopedia project. Many of them were among the first articles written for the encyclopedia, are more frequently viewed, are more frequently linked to by other articles, are longer, and more intensively edited by a greater variety of editors, than the average article. These results indicate the high value that Wikipedia users place on CS-WHL, consistent with the goal of the UNESCO World Heritage Committee to maintain a list reflects the world's cultural diversity of outstanding universal value. However, we found that sites are unequally valued, in predictable ways, given prior work on the demographics of Wikipedia users and editors, and the Eurocentric value system of the World Heritage Committee [@cleere2003uneasy; @smith2006uses p. 98]. CS-WHL in countries in the Global South are under-represented on Wikipedia, a geographical tension that is also evident in the World Heritage List [@brumann2014shifting]. They are also shorter and received less attention in the form of page views and inward Wikilinks.

Perhaps the most striking spatial pattern in our results is how much of the edits to CS-WHL articles in any part of the world come from the United States. The anonymous editors of articles about CS-WHL are generally not located in the same country as the site they are editing. This suggests a digital colonialism of world heritage information on Wikipedia, where heritage is interpreted and communicated by people who are not part of the descendant community traditionally associated with the site. This is an important observation because it is these editors that control the production of knowledge about CS-WHL articles, and the ability to control information about heritage has several implications. First is that the choices of these editors define community identity, that is, the community of Wikipedia editors, and also the broader community of Internet users that read Wikipedia, by indicating which sites are important and visible, and which are not by contributing to a digital 'authorized heritage discourse' [@smith2012constrained] that prioritizes their own self-interests. Second is that traditional communities associated with CS-WHL have little control over how their own heritage is represented on Wikipedia. We see this spatial pattern on Wikipedia as part of a trend of globalization of heritage, where places that had traditionally contributed to local or regional identities are turned into sanitized playgrounds for rich tourists [@bernbeck2004political]. Wikipedia may be seen as virtual playground for Americans to construct a universal heritage that reflects their neo-liberal cultural values, and validates their identities, perhaps reflecting a cultural rhetoric of 'heritage populism' [@reynie2016specter]. In this way, the representation of heritage on Wikipedia recapitulates a long-standing tradition of colonial subordination of one group's heritage and identity by a more powerful and dominant group.

This colonial logic may help to explain our findings of minimal conflicts surrounding the production of Wikipedia articles about CS-WHL. We found that bot activity, numbers of revert edits, and edits relating to vandalism for CS-WHL articles were comparable to or less than other sites. Most of this activity relates to playful vandalism and occurred on uncontroversial CS-WHL sites such as the Sydney Opera House and the Tower of London. Controversial sites, identified by the presence of military or diplomatic conflicts, were not prominent in these metrics. Our analysis of the talk pages of CS-WHL articles shows that they are often longer than a typical Wikipedia article, and are the location where conflict is most evidence. While much of the conflict is highly technical and local about the policies and processes of editing Wikipedia, we did observe signs of the political and military conflict at CS-WHL played out on some of the talk pages, such as Mapungubwe and Preah Vihear. We interpret this as editors exploiting the talk pages for behind-the-scenes heritage activism. The technical and cultural constraints of writing directly on the article page mean that the talk page is the only suitable location for negotiating disputes whose eventual trace in the edit history of the article might be negligible. We found that some CS-WHL talk pages referenced a 'Geopolitical, ethnic, and religious conflicts noticeboard' elsewhere on Wikipedia, as a forum for arbitration of disputes created in 2008 following a recommendation by an investigation into edit wars on articles about Northern Ireland (The Troubles), Palestine-Israel and other locations with ethnic or cultural based conflicts. However the noticeboard was closed in 2012 due to an absence of activity. Overall we find that editors of CS-WHL articles appear conflict-averse, probably because many of them come from similar cultural, socio-economic and educational backgrounds, and so share similar values about heritage that align with their shared goal of making an encyclopedia. The low levels of conflict reflect the continuity between the printed encyclopedia project as a compendium of universal knowledge, and the World Heritage List as universal heritage. These projects share origins in the cultural elites of Western European, and their distinctive sense of what is universal that is linked to nineteenth-century nationalism and liberal modernity [@smith2006uses].

The results of our time series analyses reveal complex patterns in how heritage-related concerns are realized on Wikipedia. Some sites show spikes in Wikipedia editing activity that are directly related to the inscription of the site on the WHL, or conflict events at the site. However at other sites the rhythms of editing activity seem unrelated to events directly related to the site. Like the talk pages, which have unrealised potential for close reading of heritage activism and disputes, close study individual edit histories of CS-WHL articles may reveal further patterns of how people engage with heritage on Wikipedia.

@@

The agency of people participating in heritage-related activities on Wikipedia is highly constrained by the technical characteristics of editing, bots, and consensus-valuing culture. Although there are some niches where editors can debate, concerns about state-like actors, violence & destruction, deal-making, etc. are poorly represented. Typical conflicts about Wikipedia CS-WHL articles are hyper-local, and process-centered. Spatial patterns in the edits to Wikipedia articles about CS-WHL show a kind of digital colonialism of world heritage information, with most edits emanating from the United States. Because of these technical and cultural characteristics of Wikipedia, its articles present a discourse that actively obscures the power relations that give rise to them, and makes opaque debates about how heritage is involved in the production of identity and power [cf. @harvey2001heritage].

That said, Wikipedia is valuable and important as a Big Data source of social information on heritage for several reasons. Although it lacks the velocity of other social media, such as Twitter, it preserves a distinctively complex record of human interactions with the past at simultaneously high scales, high resolution, and in highly structured ways, afforded by no other platform. As a non-profit entity, Wikipedia's data are freely available to researchers, unlike Twitter and Facebook. This has important ethical implications because it eliminates the need for academic collaboration with commercial platforms, whose business interests may not align with the researchers' ethics [@schroeder2014big]. ALthough free of these commercial interests, a different set of ethics is implied by the exquisite administrative and technical complexity of Wikipedia. These structures serve to normalize and historize inequalities, and manage threats to models of control and expertise about heritage [@harrison2015beyond]. As a context of heritage production, a resource of information on heritage, and an artefact of contemporary digital heritage, Wikipedia is remarkable and holds substantial research potential. While the Neutral Point of View is a central policy for Wikipedia articles, representations and definitions of cultural heritage are never neutral in any discourse, because it always comes from some perspective [@lahdesmaki2019conflicts]. Given the tremendous popularity and visibility of Wikipedia, is it important that future work on digital heritage in this context continue to critically examine what is absent from, and uncontested in, Wikipedia, and who benefits and suffers from these absences.

>>>>

The highly detailed edit histories that Wikipedia keeps for every article allow us to further investigate spatial and temporal questions relating to engagement with the past and conflicts surrounding CS-WHL sites. For example, when an article is edited by a user who does not have a Wikipedia user account (or is not logged into their account), their edit is identified by that person's IP address. This IP address can be used to geo-locate the user to the country they were in when the made the edit. We geo-located all edits with IP addresses for all Wikipedia articles CS-WHL sites to determine the country of origin of those edits using the rgeolocate package for R [@rgeolocate]. This helps us to answer the question: are the editors of articles about CS-WHL located near the sites they edit, indicating local community interest in the online representation of their heritage? Or are editors located in other countries, implying a digital colonialism of world heritage information on Wikipedia where one community's heritage is interpreted and communicated by another? The time and date stamps attached to every edit on every article allows us to investigate temporal patterns of activity on CS-WHL Wikipedia articles. Analyses of these data help us to answer the question: is Wikipedia editing activity correlated with events outside of Wikipedia relating to the CS-WHL sites, such as conflict events, or their inscription on the WHL?

@@

# Reproducibility and Open Source Materials

@@

Our results establish Wikipedia as an online community actively engaging with cultural heritage [@pentzold2017digging], especially cultural sites on the UNESCO World Heritage List. We have defined Wikipedia as an online peer production community with technical and cultural qualities that make it distinct from social media services such as Twitter and Facebook. On Wikipedia, the editing mechanics, bots, reverts, and talk pages work together to enable and limit discourse and debate about CS-WHL. The goal-driven culture of Wikipedia editors, and the policies that guide the production of articles strongly constrain heritage discourse, such that the exchange of different perspectives are rarely found in articles themselves. However, we have identified conflict and tension on the article talk pages, sometimes in technical and highly specific discussions, and sometimes in overtly hostile exchanges between editors. Our results show how the technicity of Wikipedia shapes how people engage with elements of the past and attribute social and cultural meanings on Wikipedia. What we have found resembles the "contingent collaborations" and "productive frictions" described by @tsing2011friction where local, individual actions of knowledge creation are circumscribed by forces and structures that encompass their expression, but also give meaning to their interactions.

Predictably, many of the Wikipedia articles about CS-WHL are central to the encyclopedia project. Many of them were among the first articles written for the encyclopedia. They are more frequently viewed, are more frequently linked to by other articles, are longer, and more intensively edited by a greater variety of editors, than the average article. These results indicate the high value that Wikipedia users place on CS-WHL, consistent with the goal of the UNESCO World Heritage Committee to maintain a list reflects the world's cultural diversity of outstanding universal value. Similarly, we found that the effort and attention sites receive on Wikipedia is highly unequal, reflecting the fairly narrow demographic traits of the majority of Wikipedia users and editors, which is further exacerbated by the Eurocentric value system of the World Heritage Committee [@cleere2003uneasy; @smith2006uses p. 98]. For example, CS-WHL in countries in the Global South are under-represented on Wikipedia, a geographical tension that is also evident in the World Heritage List [@brumann2014shifting]. Articles about CS-WHL in the Global South are also shorter and have received less attention in the form of page views and inward Wikilinks.

The most striking spatial pattern in our results is the high proportion of anonymous edits of articles about CS-WHL sites, in any part of the world, that come from the United States. These anonymous editors of articles about CS-WHL are generally not located in the same country as the site they are editing. Finding the United States at the first ranked location for anonymous edits is not surprising as it has the world's largest number of English speakers. We might similarly expect that the French language Wikipedia (2.3 m articles, 173 m edits, 3.9 m users) would have more edits originating in France or francophone on articles about CS-WHL sites in any part of the world. It is beyond the scope of this paper to analyze spatial patterns of edits in Wikipedias written in languages other than English; we focus on the English language Wikipedia because it is the largest by a considerable margin [6.2 m articles, 998 m edits, 40 m users, @wiki:listoflangs].

Although the dominance of the United States in the spatial patterns is unsurprising, our data show some patterns that that language dominance cannot not fully explain. After the United States (297 m speakers), the top five countries for numbers of English speakers are India (238 m), Nigeria (104 m), United Kingdom (60 m), Philippines (50 m), and Germany (45 m) [@Ethnologue]. However, our spatial data show that the countries that originate the largest numbers of edits to articles about foreign CS-WHL sites (ie. sites not located in the country where the edits where made) are the United States, United Kingdom, Australia and Canada. Although many edits originate from India, they are almost entirely of articles about CS-WHL sites also in India.

The United States, United Kingdom, Australia and Canada, are an important group because they are collectively known as the core Anglosphere countries, sharing a common heritage as former colonies of the British empire [\@Richards_2019]. \@vucetic2011anglosphere defines the Anglosphere as a 'distinct international, transnational, civilisational, and imperial entity within the global society. The dominance of editors of CS-WHL articles located in the Anglosphere indicates an extension of this imperial project into the organization of world heritage information on Wikipedia. The prominence of Anglosphere editors on CS-WHL articles articles represents a digital colonisation of world heritage information on Wikipedia, where heritage is interpreted and communicated by people who are not part of the descendant communities traditionally associated with the site.

This is an important observation because it is these editors that control the production of knowledge about CS-WHL articles, and the ability to control information about heritage has several implications. First is that the choices of these editors define community identity, that is, the community of Wikipedia editors, and also the broader community of Internet users that read Wikipedia, by indicating which sites are important and visible, and which are not by contributing to a digital 'authorized heritage discourse' [@smith2012constrained] that prioritizes their own self-interests (cf. the poor representation of CS-WHL sites located in the Global South). Second is that traditional communities associated with CS-WHL have little control over how their own heritage is represented on Wikipedia. We see this spatial pattern on Wikipedia as part of a trend of globalization of heritage, where places that had traditionally contributed to local or regional identities are turned into sanitized playgrounds for rich tourists [@bernbeck2004political]. Wikipedia may be seen as virtual playground for Anglosphere citizens to construct a universal heritage that reflects their neo-liberal cultural values, and validates their identities, perhaps reflecting a cultural rhetoric of 'heritage populism' [@reynie2016specter]. In this way, the representation of heritage on Wikipedia recapitulates a long-standing tradition of colonial subordination of one group's heritage and identity by a more powerful and dominant group.

This colonial logic may help to explain our findings of minimal conflicts surrounding the production of CS-WHL articles. We found that bot activity, numbers of revert edits, and edits relating to vandalism for CS-WHL articles were comparable to or less than other sites. Most of this activity relates to playful vandalism and occurred on uncontroversial CS-WHL sites such as the Sydney Opera House and the Tower of London. Controversial sites, identified by the presence of military or diplomatic conflicts, were not prominent in these metrics. Our analysis of the talk pages of CS-WHL articles shows that they are often longer than a typical Wikipedia article, and are the location where conflict is most evident. While much of the conflict is highly technical and local about the policies and processes of editing Wikipedia, we did observe signs of the political and military conflict at CS-WHL played out on some of the talk pages, such as Mapungubwe and Preah Vihear. We interpret this as editors exploiting the talk pages for behind-the-scenes heritage activism. The technical and cultural constraints of writing directly on the article page mean that the talk page is the only suitable location for negotiating disputes whose eventual trace in the edit history of the article might be negligible.

We found that some CS-WHL talk pages referenced a 'Geopolitical, ethnic, and religious conflicts noticeboard' elsewhere on Wikipedia, as a forum for arbitration of disputes created in 2008 following a recommendation by an investigation into edit wars on articles about Northern Ireland (The Troubles), Palestine-Israel and other locations with ethnic or cultural based conflicts. However the noticeboard was closed in 2012 due to an absence of activity. Overall we find that editors of CS-WHL articles appear conflict-averse, probably because many of them come from similar cultural, socio-economic and educational backgrounds due to their Anglosphere origins. This means they share similar values about heritage that align with their shared goal of making an encyclopedia. The low levels of conflict reflect the continuity between the printed encyclopedia project as a compendium of universal knowledge, and the World Heritage List as universal heritage. These projects share origins in the cultural elites of Western Europe, and their distinctive sense of what is universal that is linked to nineteenth-century nationalism and liberal modernity [@smith2006uses].

The results of our time series analyses reveal complex patterns in how heritage-related concerns are realized on Wikipedia. Some sites show spikes in Wikipedia editing activity that are directly related to the inscription of the site on the WHL, or conflict events at the site. However at other sites the rhythms of editing activity seem unrelated to events directly related to the site. Like the talk pages, which have potential for close reading of heritage activism and disputes, close study of individual edit histories of CS-WHL articles may reveal further patterns of how people engage with heritage on Wikipedia.

@@

The agency of people participating in heritage-related activities on Wikipedia is highly constrained by the technical characteristics of editing, bots, and consensus-valuing culture. Although there are some niches where editors can debate, concerns about state-like actors, violence & destruction, deal-making, etc. are minor. Typical conflicts about Wikipedia CS-WHL articles are hyper-local, and process-centered, but there are also cases of editorial conflicts about substantive issues of specific CS-WHL. Spatial patterns in the edits to Wikipedia articles about CS-WHL show a kind of digital colonialism of world heritage information, with most edits emanating from the Anglosphere, and poor coverage of CS-WHL in the Global South. Because of these technical and cultural characteristics of Wikipedia, its articles present a discourse that actively obscures the power relations that give rise to them, and makes opaque debates about how heritage is involved in the production of identity and power [cf. @harvey2001heritage].

That said, Wikipedia is valuable and important as a Big Data source of social information on heritage for several reasons. Although it lacks the velocity of other social media, such as Twitter, it preserves a distinctively complex record of human interactions with the past at simultaneously high scales, high resolution, and in highly structured ways, afforded by no other platform. As a non-profit entity, Wikipedia's data are freely available to researchers, unlike Twitter and Facebook. This has important ethical implications because it eliminates the need for academic collaboration with commercial platforms, whose business interests may not align with the researchers' ethics [@schroeder2014big]. Although free of these commercial interests, a different set of ethics is implied by the exquisite administrative and technical complexity of Wikipedia. These structures serve to normalize and historize inequalities, and manage threats to models of control and expertise about heritage [@harrison2015beyond]. As a context of heritage production, a resource of information on heritage, and an artefact of contemporary digital heritage, Wikipedia is remarkable and holds substantial research potential. While the Neutral Point of View is a central policy for Wikipedia articles, representations and definitions of cultural heritage are never neutral in any discourse, because they always come from some perspective [@lahdesmaki2019conflicts]. Given the tremendous popularity and visibility of Wikipedia, is it important that future work on digital heritage in this context continue to critically examine what is absent from, and uncontested in, Wikipedia, and who benefits and suffers from these absences.
